One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2024-10-31 paper.tex
intro edits

Text width here we go:
%\printinunitsof{cm}\prntlen{\textwidth}

The estimation of the size and spatial distribution of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}. Here we present an analysis of wildlife survey data collected on a critically endangered Hawaiian forest bird that aims to meet this objective.  The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.  The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution.  This information is critical when decisions about conservation strategies are made.  However, estimating the abundance and spatial distribution of wild populations of animals presents many statistical challenges.

First, as in many ecological surveys, it is impossible to undertake a full census (i.e. the complete enumeration of all individuals within a defined study region).  The existing population numbers in the thousands and lives in dense tropical forest where logistical and ecological challenges mean a census is not feasible.  For this reason, a monitoring survey must sub-sample appropriately in space and time.  For the \akepa{} this takes the form of the \hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}, which is a large-scale, quantitative survey of Hawaiian forest birds.  Annual surveys of the \akepa{} study-region consist of a number of point transects located along randomly located line transects.

Second, even at surveyed locations, the detectability of animals is unknown.  The \akepa{} survey estimates detectability using a point transect distance sampling approach \citep{buckland_distance_2015} where, for each observation, the distance to the observer is recorded.  The probability of detecting a bird is modelled using a parametric detection function that decays with increasing distance.  The detection function parameters are estimated from the observed distances by assuming the true density is uniform with respect to distance and attributing deviations from uniformity to imperfect detectability. The combination of spatio-temporal subsampling and unknown detectability means statistical methods to analyse distance sampling data must model a complex observation process alongside the spatial distribution of animals.

Third, a key aim of the analysis is to predict animal density at unsurveyed locations.  If animal density is modelled using spatial covariates then the covariate effects estimated in the surveyed region can be extrapolated to unsurveyed location.  However, in many ecological contexts there can be good reason to believe there are drivers of the spatial distribution for which we have no explanatory covariates available or for which, in principle, no covariate could be constructed (e.g. the `sociality' of the species).  For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient for which no straightforward explanatory cause has been found \citep{camp_dsm_2020}.  From a statistical perspective, this heterogeneity that cannot be explained by available covariates suggests the inclusion of spatially structured random effects in the model.  Such effects require a careful consideration of the model structure and inference methods to ensure that model fitting remains computationally feasible.

Lastly, the above challenges mean the resulting statistical model is necessarily complex. It is therefore challenging to communicate the results of the analysis to non-statistically trained audiences.  The \akepa{} data are collected with the clear objective of monitoring the population and informing conservation management decisions.  These decisions will be taken by stakeholders with a range of statistical expertise.  This final step of communicating results should therefore be considered with as much care as the rest of the analysis.  This requires the input of statisticians to ensure uncertainty is appropriately captured and communicated in the key outputs of the analysis upon which decisions will be based.  In particular, in the context of species distribution modelling, we note the challenge of communicating uncertainty in maps of predicted animal density.

We present an analysis that addresses these statistical challenges by using the following approaches:

\begin{enumerate}[(a)]
	\item A spatial point process perspective on point transect distance sampling and representing the detection model as a thinning of a point processes.
	\item A one-stage approximate Bayesian approach to inference based on repeated model fits using integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009}.  This allows for the simultaneous estimation of the spatial distribution and detectability of animals.
	\item A computationally efficient Gaussian Markov random field (GMRF) spatially structured random effect specified through a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011} to account for unobserved drivers of the spatial distribution.
	\item A discussion on model evaluation and communication of results that takes advantage of the one-stage approach by sampling from the joint posterior of all model parameters. We advocate for excursions methods \citep{bolin_excursion_2015} as a way to handle uncertainty in spatial predictions of density.  
\end{enumerate}

We demonstrate several novel contributions to the problem of species distribution modelling under unknown detectability.  This is, to the best of our knowledge, the first analysis of point transect distance sampling data using a point process perspective.  This requires a modified intensity function that accounts for incomplete location information for animal sightings.  This is also the first application of a novel approximate Bayesian inference method to estimate the detection function jointly with the spatial model.  This inference method is readily generalisable to other applications beyond distance sampling models.  As such, this analysis will be of interest to the broad applied statistics community in general, as well as those specifically interested in modelling species distributions.

The rest of the paper proceeds as follows:  Section \ref{sec-study-design} describes the \akepa{} study region and survey design; Section \ref{sec-ds-pp} reviews current approaches to distance sampling and presents the point process perspective; Sections \ref{sec-iinla} and \ref{sec-gmrf} describe the iterated INLA fitting procedure and the GMRF random effect; and Sections \ref{sec-results} presents results of the analysis and discusses model evaluation and effective communication of uncertainty.

>>>>

The estimation of the abundance and spatial distribution of wild populations of animals is a critical objective in ecology and conservation \citep{schwarz_estimating_1999}.This paper presents a statistical analysis of survey data collected on a critically endangered Hawaiian forest bird, offering insights into its population size and spatial dynamics, which crucial to guide effective conservation management decisions.  The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.  The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution through time.

However, estimating the abundance and spatial distribution of wild populations of animals presents various statistical challenges. Traditional methods are usually split into two stages, where the spatial distribution is estimated in a second-stage spatial model, conditional on estimates of the probability of detecting animals given they are present.  This detectability is estimated in a first-stage distance sampling model \citep{millerExtendingDensitySurface2021, miller_spatial_2013}.  The two-stage approach requires either the loss of uncertainty from the first stage, or the use of approximate methods to propagate uncertainty between the modelling stages \citep{bravington_VariancePropagationDensity_2021}.  In this paper, we present the first one-stage analysis of point transect distance sampling data that simultaneously estimates the non-linear detectability and spatial distribution of the \akepa{} population.  In order to achieve this, we take a point process perspective on distance sampling survey data which requires the use of a novel approach to inference that allows estimation of non-linear parametric model components within a generalized additive model framework. This approach is based on an iterative scheme of successive fits using integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009} that is implemented in the software package \texttt{inlabru} \citep{lindgren_inlabru_2024}.  This approach is flexible, general and applicable in a wide range of contexts beyond modelling species distributions.

Since the existing \akepa{} population numbers in the thousands and lives in dense tropical forest where logistical and ecological challenges mean a full census of the population is not feasible. The \hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986} is a large-scale, quantitative survey of Hawaiian forest birds.  Annual surveys of the \akepa{} study-region consist of a number of point transects located at approximately regular intervals along randomly located line transects.  At these surveyed locations, the detectability of animals is unknown and must be estimated in order to convert observed counts into estimates of the total population size.  The \akepa{} survey estimates detectability using a point transect distance sampling approach \citep{buckland_distance_2015} where, for each observation, the distance to the observer is recorded.

Additionally, a key aim of the analysis is to predict animal density at unsurveyed locations.  If animal density is modelled using spatial covariates then the covariate effects estimated in the surveyed region can be extrapolated to unsurveyed location.  However, in many ecological contexts there is good reason to believe there are drivers of the spatial distribution for which we have no suitable explanatory covariates available. For the \akepa{} data, there is a clear example of an unexplained driver of the spatial distribution where the population favours the southern region of the preserve.  No straightforward explanatory cause has been found for this preference \citep{camp_dsm_2020}.  From a statistical perspective, we address this unexplained heterogeneity by including a spatially structured random effect in the model.

These challenges in analysing distance sampling survey data mean that the resulting statistical model is necessarily complex. Since the \akepa{} data are collected with the clear objective of monitoring the population and informing conservation management decisions, there is a challenge for statisticians to effectively communicate with stakeholders who have a wide range of statistical expertise. We believe the Bayesian one-stage approach is a useful step to support statisticians in this task.  Since model outputs are based on Monte Carlo samples from the joint posterior, summary outputs naturally marginalise over the uncertainty in the observation process in a way that would be challenging in the traditional maximum-likelihood two-stage context.  We discuss various challenges associated with communicating uncertainty in mapped estimates of animal density and propose some illustrative alternatives that make use of the Bayesian one-stage perspective.

The paper is structured as follows:  Section \ref{sec-study-design} describes the \akepa{} study region and survey design; Section \ref{sec-ds-pp} reviews current approaches to distance sampling and presents the point process perspective; Sections \ref{sec-iinla} and \ref{sec-gmrf} describe the iterated INLA fitting procedure and the spatially-structured random effect; and Sections \ref{sec-results} presents results of the analysis, model evaluation and effective communication of uncertainty.
